---
title: 'Repeated Measures'
#date: '`r format(Sys.time(), "%d %B %Y")`'
bibliography: module_refs.bib
csl: https://www.zotero.org/styles/harvard-cite-them-right
output: 
   html_document:
    toc: true
    toc_depth: 3 
---
---

```{r setup}
library(tidyverse)
library(gridExtra)
library(emmeans) # for emmeans() and contrast()
options(width=100)
```

---

# Objectives

* Understand the analysis of repeated measures data
* Understand the difference between _null hypothesis significance testing_ (NHST) and _estimation_
* Be able to run repeated measures analysis in R:
	* Using `aov` and `summary`, for the NHST approach
	* Using `emmeans`, for the estimation approach
* Be able to report results in plain English using a NHST or estimation approach
* Be able to produce excellent plots with `ggplot`

---

# Reading

@Howell17 Chapters 13 and 18 are an excellent introduction to repeated measures

---

# A workplace well-being intervention

We measured workers' well-being every week for five weeks. Between weeks 2 and 3, we make a workplace intervention intended to increase well-being

```{r load data}
intervention <- read_csv("well-being.csv")
intervention <- intervention %>% mutate(time=as.factor(time))
intervention <- arrange(intervention, sub, time)
intervention
```

---

# The incorrect ANOVA

```{r plot data wrongly}
ggplot(intervention, aes(x=time, y=well.being)) + geom_jitter(height=0.05, width=0.1) + scale_y_continuous(breaks=1:10) + labs(x="Time / Weeks", y="Well-Being", subtitle="The dotted line marks the intervention") + geom_vline(xintercept=2.5, lty=3)
```

```{r incorrect model}
m.incorrect <- lm(well.being~time, data=intervention)
summary(m.incorrect)
anova(m.incorrect)
```

According to this _incorrect_ ANOVA, there is no significant effect of time, $F(4,35)=2.1$, $p=.11$. 

---

## Contrasts (on the incorrect model)

We can hand-build some custom contrasts

`after.minus.before` compares the average of times 1 and 2 with the average of times 3, 4, and 5

`t3.minus.t2` compares times 2 and 3

```{r incorrect contrasts}
(  m.incorrect.emm <- emmeans(m.incorrect, ~time)  )
(  contrasts <- data.frame(after.minus.before=c(-1/2, -1/2, 1/3, 1/3, 1/3), t3.minus.t2=c(0, -1, 1, 0, 0))  )
confint(contrast(m.incorrect.emm, contrasts))
```

According to this _incorrect_ ANOVA, the contrast for before and after the treatment shows that well-being increases by 1.67 points 95% CI[0.43--2.90]

---

# The correct repeated-measures ANOVA

In repeated measures data, observations are not considered independent draws from the population. We are taking repeated observations from the same unit of analysis (e.g., the same person, the same company)

_Always_ plot repeated measures data _for each participant_

```{r good plotting}
ggplot(intervention, aes(x=time, y=well.being, group=sub, col=sub)) + geom_line(alpha=0.5) + scale_y_continuous(breaks=1:10) + labs(x="Time / Weeks", y="Well-Being", col="Subject") + geom_vline(xintercept=2.5, lty=3)
```

The `aov()` function is used for repeated measures data. Note the extra `Error(sub/time)` term, which means that time is varying within each subject---that each subject is measured at each time

The code `contrasts(intervention$time) <- contr.sum` does something technical. Because there is no natural time point to call out reference level, we are using _effects coding_ (also called _sum to zero contrasts_). This makes sense whenever there is no natural reference level 

```{r repeated measures ANOVA}
#contrasts(intervention$time) <- contr.treatment # The default contrast, where each level is contrasted with the first
contrasts(intervention$time) <- contr.sum # Each level is contrasted with the grand mean
m.well.being.by.time <- aov(well.being~time + Error(sub/time), data=intervention)
summary(m.well.being.by.time)
```

We write: "The effect of time is significant, $F(4,28)=21.4$, $p<.0001$"

## Contrasts on the correct repeated-measures ANOVA

```{r contrasts on repeated measures ANOVA}
(  m.well.being.by.time.emm <- emmeans(m.well.being.by.time, ~time)  )

(  contrasts <- data.frame(after.minus.before=c(-1/2, -1/2, 1/3, 1/3, 1/3), t3.minus.t2=c(0, -1, 1, 0, 0))  )
confint(contrast(m.well.being.by.time.emm, contrasts))
```

"Well-being is 1.7 points higher 95% CI[1.3--2.1]  after the intervention"

---

## Adjusted well-being

Averaging over weeks, some employees just have higher well-being than others. These individual differences are hiding the effect of our intervention

```{r well-being with average}
(  mean.well.being <- intervention %>% group_by(sub) %>% summarise(mean.well.being=mean(well.being))  )

ggplot(intervention, aes(x=time, y=well.being, group=sub, col=sub)) + geom_line() + facet_grid(.~sub) + geom_hline(data=mean.well.being, aes(yintercept=mean.well.being, col=sub), lty=2) + geom_vline(xintercept=2.5, lty=3) + labs(x="Time / Weeks", y="Well-Being", col=NULL) + scale_y_continuous(breaks=1:10)
```

Let's adjust each person's well-being score by subtracting their own personal mean

```{r adjusted well-being}

intervention <- merge(intervention, mean.well.being)
intervention <- intervention %>% mutate(adjusted.well.being=well.being-mean.well.being)

ggplot(intervention, aes(x=time, y=adjusted.well.being, group=sub, col=sub)) + geom_line(alpha=0.7) + labs(x="Time / Weeks", y="Deviation from Average Well-Being", col="Subject") + geom_vline(xintercept=2.5, lty=3) + ylim(-5,5)
```

---

## Sources of variability

In the repeated-measures ANOVA all the variability due to subject differences is removed when thinking about the effect of time.

```{r sources of variability, echo=FALSE}
grid.arrange(
	ggplot(intervention, aes(x=time, y=well.being, group=sub)) + geom_point(alpha=0.5) + labs(x="Time / Weeks", y="Average Well-Being", title="With Between-Subject Variability") + geom_vline(xintercept=2.5, lty=3) + ylim(0,10),
	ggplot(intervention, aes(x=time, y=adjusted.well.being, group=sub)) + geom_point(alpha=0.5) + labs(x="Time / Weeks", y="Deviation from Average Well-Being", title="Without Between-Subject Variability") + geom_vline(xintercept=2.5, lty=3) + ylim(-5,5),
	ncol=2
)
```

```{r sum of squares, echo=FALSE}
# I do not expect you to ever run this code, so I've hidden it
incorrect.anova.SS <- c(0, anova(m.incorrect)$"Sum Sq")
names(incorrect.anova.SS) <- c("Subjects", "Time", "Residuals")
correct.repeated.measures.anova.SS <- c((summary(m.well.being.by.time)$"Error: sub")[[1]]$"Sum Sq", (summary(m.well.being.by.time)$"Error: sub:time")[[1]]$"Sum Sq")
names(correct.repeated.measures.anova.SS) <- c("Subjects", "Time", "Residuals")
SS <- data.frame(model=rep(c("Incorrect ANOVA", "Repeated Measures ANOVA"), each=3), SS=c(incorrect.anova.SS, correct.repeated.measures.anova.SS), source=c(names(incorrect.anova.SS), names(correct.repeated.measures.anova.SS)))
ggplot(SS, aes(x=source, y=SS)) + geom_col() + facet_grid(.~model) + labs(x="Source", y="Variability")
```

---

## Paired-samples $t$-test

Let's compare before vs after with a paired samples $t$-test. This is nearly the same as the before--after contrast

```{r before after t-test}
intervention <- intervention %>% mutate(timing=ifelse(time %in% 1:2, "Before", "After"))
(  intervention.timing <- intervention %>% group_by(sub, timing) %>% summarise(well.being=mean(well.being))  )
(  intervention.timing.wide <- intervention.timing %>% spread(timing, well.being)  )
t.test(intervention.timing.wide$Before, intervention.timing.wide$After, paired=TRUE)
```

---

# Mixed ANOVA


_Always_ plot repeated measures data _for each participant_

```{r mixed ANOVA plotting}
ggplot(intervention, aes(x=time, y=well.being, group=sub, col=sex)) + geom_line(alpha=0.5) + scale_y_continuous(breaks=1:10) + labs(x="Time / Weeks", y="Well-Being", col="Sex", subtitle="Each line is one person") + geom_vline(xintercept=2.5, lty=3)
```

```{r mixed ANOVA}
m.well.being.by.time.sex <- aov(well.being~time*sex + Error(sub/time), data=intervention)
summary(m.well.being.by.time.sex)
```

We write: "The effect of time is significant, $F(4,24)=19.7$, $p<.0001$. The effect of time does not interact with sex, $F(4,23)=0.7$, $p=.8$"

## Contrasts on the correct repeated-measures ANOVA

```{r contrasts on mixed ANOVA}
(  m.well.being.by.time.sex.emm <- emmeans(m.well.being.by.time.sex, ~time + sex)  )

(  interaction.contrast <- data.frame(timing.by.sex.interaction=c(-1/2, -1/2, 1/3, 1/3, 1/3, 1/2, 1/2, -1/3, -1/3, -1/3))  )
confint(contrast(m.well.being.by.time.sex.emm, interaction.contrast))
```

"The improvement in well-being for men is 0.17 95% CI[-0.64--0.98] smaller than the improvement for women"

# References

