---
title: 'Gender Pay Gap: $t$-test workshop '
author: Neil Stewart
#date: '`r format(Sys.time(), "%d %B %Y")`'
bibliography: module_refs.bib
csl: https://www.zotero.org/styles/harvard-cite-them-right
output: 
   html_document:
    toc: true
    toc_depth: 3 
---
---

```{r setup}
library(tidyverse)
library(lubridate)
options(width=100)
```

---

# Week 4 Workshop

In this workshop we will run t-tests on the gender pay gap data

When you run your $t$-tests, be sure to write plain English sentences to explain the results

Be sure to include

* Describe in words what you want the reader to know
* Don't forget to describe the _difference_ of the direction (so "The heights of boys differ significantly from the heights of girls" is bad, but "The boys were significantly shorter than the girls" is good)
* Report means
* Report the $t$ statistic, the degrees of freedom, and the $p$ value

You will need to combine your code from the 'dplyr' and 'ggplot' workshops

## Tasks for today

1. Make a sample of 100 fake data points drawn from a normally-distributed population mean 10 and standard deviation 10. Run a $t$-test on your sample testing the hypothesis that the population mean is zero
	* What happens when you change the mean to 20? 
	* What happens when you change the standard deviation to 5?
	* What happens when you change the sample size to 500?
For each question, explain _why_

(Hint: `rnorm(n=100, mean=10, sd=10)` will sample 100 random observations from a normally-distributed population mean 10 and standard deviation 10)

2. Is the gender gap in mean hourly pay zero?

3. Are public sector employers better than businesses or charities? 

4. Is the pay gap worse for employers submitting late? 

For each question:

* Write out more carefully what the hypothesis the $t$-test is testing here actually is
* Run the $t$-test
* Write a sentence describing the inference you can make. Include the $t$-test results

Finally, check you can knit your Rmarkdown document to produce an HTML file documenting everything you have done

## Learning objectives for today

1. Understand the $t$-test
2. Learn how to run $t$-tests in R
3. Learn how to report $t$-tests
  
---

# Sampling: Question 1

```{r}
t.test(rnorm(n=100, mean=10, sd=10))
t.test(rnorm(n=100, mean=20, sd=10))
t.test(rnorm(n=100, mean=10, sd=5))
t.test(rnorm(n=500, mean=10, sd=10))
```

When the sample mean increases, the standard deviation decreases, or the sample size increases, $t$ will get larger and $p$ will get smaller. We are more sure that the population mean is not zero when the sample mean increases, the standard deviation decreases, or the sample size increases

# Gender Pay Gap

```{r}
pay.gap <- read_csv("UK_gender_pay_gap_2017-18.csv")
```

---

## Question 2: Is the gender gap in mean hourly pay zero?

```{r}
pay.gap %>% summarise(mean=mean(DiffMeanHourlyPercent), N=n())
ggplot(pay.gap, aes(x=DiffMeanHourlyPercent)) + geom_histogram(binwidth=1) + xlim(-100, 100) + labs(x=expression(frac(Male~Mean-Female~Mean,Male~Mean)*'%'), y="Frequency", subtitle="NB 5 observations <100 omitted and 1 observation>100 omitted")
```

Figure X. The distribution of gender pay gap. Pay gap is defined as the difference between mean male pay and mean female pay as a fraction of mean male pay. Bars are bins 1% wide

```{r}
t.test(pay.gap$DiffMeanHourlyPercent)
```

The pay gap in mean hourly pay 14.3%, which is significantly above zero, $t(10531)=96.5$, $p<.0001$

---

# Question 3: Are public sector employers better than businesses or charities?

```{r}
pay.gap <- mutate(pay.gap, DueDate=dmy_hms(DueDate), Sector=ifelse(DueDate==ymd("2018-03-31"), "Public Sector", "Business / Charity"))

(  pay.gap.by.sector <- pay.gap %>% group_by(Sector) %>% summarise(mean=mean(DiffMeanHourlyPercent), n=n())  )

pay.gap.by.sector %>% summarise(difference=diff(mean))


ggplot(pay.gap) + geom_histogram(aes(x=DiffMeanHourlyPercent), binwidth=1) + facet_grid(Sector~.) + xlim(-100,100) + labs(x=expression(frac(Male~Mean-Female~Mean,Male~Mean)*'%'), y="Frequency", subtitle="NB 5 observations <100 omitted and 1   observation>100 omitted") + geom_vline(data=pay.gap.by.sector, mapping=aes(xintercept=mean), col="green")
```

Figure X. The distribution of gender pay gap by industry sector. Green lines mark distribution means

```{r}
t.test(DiffMeanHourlyPercent~Sector, data=pay.gap)
```

Business / charity employers have a mean hourly pay gap of 14.2%. Public sector employers have a higher mean hourly pay gap of 14.9%. This difference of 0.7% is significant, $t(3304.5)=2.40$, $p=.016$

---

# Question 4: Does submitting late predict pay gap?

```{r}
pay.gap <- pay.gap %>% mutate(SubmittedAfterTheDeadline=factor(SubmittedAfterTheDeadline, levels=c("True", "False"), labels=c("Late Submissions", "On-Time Submissions")))

(  pay.gap.by.late <- pay.gap %>% group_by(SubmittedAfterTheDeadline) %>% summarise(frequency=n(), mean=mean(DiffMeanHourlyPercent))  )

pay.gap.by.late %>% summarise(difference=diff(mean))

ggplot(pay.gap, aes(x=DiffMeanHourlyPercent, y=..density..)) + geom_histogram(binwidth=1) + facet_grid(SubmittedAfterTheDeadline~.) + xlim(-100,100) + labs(x=expression(frac(Male~Mean-Female~Mean,Male~Mean)*'%'), y="Density", subtitle="NB 5 observations <100 omitted and 1 observation>100 omitted") + geom_density()
```

Figure X. The distribution of gender pay gap by submission time. Bars show the density in 1% bins, and the black line shows the Gaussian smoothed density using Silverman's rule of thumb

```{r}
t.test(DiffMeanHourlyPercent~SubmittedAfterTheDeadline, data=pay.gap)
```

Employers submitting on time have a pay gap of 14.5%. Employers submitting late have a pay gap of 11.2%, which is a significant 3.3% lower, $t(705.56)=4.69$, $p<.0001$


