---
title: "K-means_Cluster_Analysis"
author: "Ingeon Hwang"
date: "02/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means Cluster Analysis

Clustering is a broad set of techniques for finding subgroups of observations within a data set. When we cluster observations, we want observations in the same group to be similar and observations in different groups to be dissimilar.

Process 
1. Replication Requirements: What you’ll need to reproduce the analysis 
2. Data Preparation: Preparing the data for cluster analysis
3. Clustering Distance Measures: Understanding how to measure differences in observations
4. K-Means Clustering: Calculations and methods for creating K subgroups of the data
5. Determining Optimal Clusters: Identifying the right number of clusters to group the data

```{r}
library(readxl)
library(tidyverse)  # data manipulation
library(cluster)    # clustering algorithms
library(factoextra) # clustering algorithms & visualization
```


#### Data Preparation
```{r}
wage_data <- read_excel("wagessurvey.xlsx")
wage_df <- as.data.frame(wage_data)

wage_selected<- select(wage_df, Education, Experience, Age, Wage)

wage_selected <- na.omit(wage_selected) # To remove any missing value that might be present in the data

wage_selected <- scale(wage_selected)
head(wage_selected)

```

```{r}
distance <- get_dist(wage_selected)
fviz_dist(distance, gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07"))
```
It classifies objects in multiple groups (i.e., clusters), such that objects within the same cluster are as similar as possible (i.e., high intra-class similarity), whereas objects from different clusters are as dissimilar as possible (i.e., low inter-class similarity). In k-means clustering, each cluster is represented by its center (i.e, centroid) which corresponds to the mean of points assigned to the cluster.

```{r}
wage_kmean <- kmeans(wage_selected, centers = 2, nstart = 25)
str(wage_kmean)
wage_kmean
```
The first step when using k-means clustering is to indicate the number of clusters (k) that will be generated in the final solution

K-means algorithm can be summarized as follows:

Specify the number of clusters (K) to be created (by the analyst)
Select randomly k objects from the data set as the initial cluster centers or means
Assigns each observation to their closest centroid, based on the Euclidean distance between the object and the centroid
For each of the k clusters update the cluster centroid by calculating the new mean values of all the data points in the cluster. The centroid of a Kth cluster is a vector of length p containing the means of all variables for the observations in the kth cluster; p is the number of variables.
Iteratively minimize the total within sum of square (Eq. 7). That is, iterate steps 3 and 4 until the cluster assignments stop changing or the maximum number of iterations is reached. By default, the R software uses 10 as the default value for the maximum number of iterations.

```{r}
fviz_cluster(wage_kmean, data = wage_selected)
```
We can also view our results by using fviz_cluster. This provides a nice illustration of the clusters. If there are more than two dimensions (variables) fviz_cluster will perform principal component analysis (PCA) and plot the data points according to the first two principal components that explain the majority of the variance.

```{r}
k2 <- kmeans(wage_selected, centers = 2, nstart = 25)
k3 <- kmeans(wage_selected, centers = 3, nstart = 25)
k4 <- kmeans(wage_selected, centers = 4, nstart = 25)
k5 <- kmeans(wage_selected, centers = 5, nstart = 25)

# plots to compare
p1 <- fviz_cluster(k2, geom = "point", data = wage_df) + ggtitle("k = 2")
p2 <- fviz_cluster(k3, geom = "point",  data = wage_df) + ggtitle("k = 3")
p3 <- fviz_cluster(k4, geom = "point",  data = wage_df) + ggtitle("k = 4")
p4 <- fviz_cluster(k5, geom = "point",  data = wage_df) + ggtitle("k = 5")

library(gridExtra)
grid.arrange(p1, p2, p3, p4, nrow = 2)
```

```{r}
set.seed(123)

# function to compute total within-cluster sum of square 
wss <- function(k) {
  kmeans(wage_selected, k, nstart = 10 )$tot.withinss
}

# Compute and plot wss for k = 1 to k = 15
k.values <- 1:15

# extract wss for 2-15 clusters
wss_values <- map_dbl(k.values, wss)

plot(k.values, wss_values,
       type="b", pch = 19, frame = FALSE, 
       xlab="Number of clusters K",
       ylab="Total within-clusters sum of squares")
```

```{r}
set.seed(123)

fviz_nbclust(wage_selected, kmeans, method = "wss")
```

```{r}
# function to compute average silhouette for k clusters
avg_sil <- function(k) {
  km.res <- kmeans(wage_selected, centers = k, nstart = 25)
  ss <- silhouette(km.res$cluster, dist(df))
  mean(ss[, 3])
}
```

```{r}
fviz_nbclust(wage_selected, kmeans, method = "silhouette")
```

```{r}
# compute gap statistic
set.seed(123)
gap_stat <- clusGap(wage_selected, FUN = kmeans, nstart = 25,
                    K.max = 10, B = 50)
# Print the result
print(gap_stat, method = "firstmax")
```

```{r}
fviz_gap_stat(gap_stat)
```

```{r}
# Compute k-means clustering with k = 4
set.seed(123)
final <- kmeans(wage_selected, 3, nstart = 25)
print(final)
```

```{r}
fviz_cluster(final, data = wage_selected)
```

